Local PCA Shows How Population Structure Differs Along the Genome

### 2016-06-23 writeup/local_pca_paper.tex
edited partway through results

In all these 3 species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
This implies that variation is truly due to population structure, 
since fluctuations due to demographic noise are not expected to show long distance correlations. 
Below, we display the results and investigate likely underlying causes.

\subsection{\textit{Drosophila melanogaster}}
We ran the above method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
For each, the two-dimensional MDS visualization resembles a triangle. (eg. Figure \ref{fig:mds_chr2L}a) 
Since the relative position of each window in this plot shows the similarity between windows, this suggests that there are 3 extreme types of population structure shown in the 3 peaks of the ``triangle'', and that other window's population structure might be a mixture of those extremes.

\begin{figure}
    \begin{center}
       \includegraphics{Fig1_allchr_Together_MDS_plot_compact}
    \end{center}
    \caption{
         Variation in structure for windows on chromosome arms 2L, 2R, 3L, 3R and X. 
         Each represents a window: the first column shows the MDS visualization of relationships between windows. 
         The second and third columns show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
         Rows show results from different chromosome arms, from top to bottom: Chr2L, Chr2R, Chr3L, Chr3R, ChrX. 
         Colors are consistent for plots in each row. 
         Vertical black lines show the breakpoints of known polymorphic inversions for Chr2L, Chr2R, Chr3L, Chr3R (from top to down: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
         \label{fig:mds_chr2L}
    }
\end{figure}

To investigate these extremes, we pick a window for each extreme, 
and take out the 5\% of windows that are closest to it in the MDS coordinates, 
then combine those windows for each extreme and apply PCA on the corresponding sections of genome. 
We can see in Figure \ref{fig:pca_chr2L} the obvious difference between their PCA plots.

\begin{figure}
    \begin{center}
       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L}. 
        Each point represents a sample. 
        The plots are plotting the second principal components against the first principal components. 
        The first column shows the windows colored green in Figure 1; the second column shows orange windows; and third shows purple windows.
         \label{fig:pca_chr2L}
    }
\end{figure}

There's a large inversion on Chr2L in that is polymorphic in these samples, In(2L)t \citep{corbett2012population}. We recolored the PCA plots in Figure \ref{fig:pca_chr2L} by the orientation of the inversion for each sample. Figure \ref{fig:color_inver}
Two regions of similar, extreme population structure (green in Figure \ref{fig:mds_chr2L}) are found around inversion breakpoints, 
and the other two extremes occur in the center of the inversion and between the inversion and centromere.
The corresponding PCA plots show that locally, population structure is mostly determined by which orientation of the inversion each sample has.
Similar results are found in other chromosome arms that have known polymorphic inversions (Chr2R, Chr3L, Chr3R) in \textit{Drosophila}. 
Other known polymorphic inversions are like In(3L)P on Chr3L, In(3R)Mo and  In(3R)P on Chr3R. 
In(3L)P and In(3R)Mo are just a few in our samples, thus we don't see the significant influence of population structure of them. 
The situation of Chr3R is a little complicated. The coexisting inversions make the MDS visualiazation of it more difficult to pick the extremes.
For the extremes we picked in Figure \ref{fig:mds_chr2L}, In(3R)K could best explain the PCA clustering result, 
while picking other windows as extremes might show that In(3R)P explain the PCA clustering result.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversion, In(2L)t, In(2R)NS, In(3L)OK and In(3R)K. (Data from \citet{lack2015drosophila}.
        \label{fig:color_inver}
    }
\end{figure}

\subsection{Human}
We ran our method separately on all 22 human autosomes. 
For instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) coincide with the position of known polymorphic inversions on 8p23. 
Similar results are found in other chromosomes that have known inversions (eg. Chromosome 15, Chromosome 17).
We found that the primary axis of variation in population structure differentiated only one or a few windows on each chromosome, 
unlike the countinous variation in population structure we see in Drosophila.
Other chromosomes showed similar results around predicted inversions: PCA might provide an additional way to identify inversions \citep{ma2012investigation}.

When we run the method on all 22 autosomes together, the outlying signal of chromosome 8 is still visible. (See supplementary for all 22 autosomes)

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
    \end{center}
    \caption{
         Variation in structure between windows on human chromosomes 8, 15, and 17. 
         Each point in the plot represents a window.
         The first column shows the MDS visualization of relationships between windows;
         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
         Rows, from top to bottom show chromosomes 8, 15, and 17. 
         The vertical red lines show the breakpoints of the known inversions \citep{antonacci2009characterization}.
        \label{fig:mds_human}
    }
\end{figure}

@@

\plr{Check if I've missed anything, then remove these?}

\subsection{Future work}
1. For human and Drosophila, we want to eliminate the regions under known inversions and check the variation of population structure for the remaining part by removing those sections. We try to check whether they will give similar results as in \textit{Medicago truncatula}, that is whether the variation is closely related to heterochromatin or gene density.

\noindent 2. Uneven sampling has a strong influence on PCA projections \citep{mcvean2009genealogical}. 
Our human data, POPRES, is unevenly sampled including 346 African-Americas, 73 Asians, 359 Indian Asians and 3187 Europeans. 
First, we'll try sub-sampling Europeans to balance the population size for the 4 population and repeat the process on the resampled data. 
Second, we'll try to apply the whole process on only European samples to see the genetic variation inside European samples. 
Third, we want to try different scheme of adding a weighting matrix to the covariance matrix of genotype data, thus to the reduce the influence of uneven sampling.

\noindent 3. Since regions that have low recombination rate tend to have similar PCs, we'll try cutting the chromosomes into windows with same distance in genetic map instead of same SNP numbers.

\noindent 4. Euclidean distance between the contracted matrix based on PCs is one measure of the similarity for window's population structure.
We want to try other methods of distance between windows, for example, we used the distance for PCs to reduce noise, however the distance between covariance matrixes of genotype matrix might also be informative.

\noindent 5. Although the first two coordinates contains the main part of information, we'd like to see the information contained in higher PCs (e.g. the third PC, the forth PC), and higher dimension of MDS.

>>>>

In these three species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
This implies that variation is due to meaningful variation in population structure, 
since fluctuations due to demographic noise are not expected to show long distance correlations. 
Below, we discuss the results and likely underlying causes.

%%%%%%%%%%%
\subsection{\textit{Drosophila melanogaster}}

We ran the method on chromosome arms 2L, 2R, 3L, 3R and X separately. 
For each, the two-dimensional MDS visualization resembles a triangle,
as seen in Figure \ref{fig:mds_chr2L}.
Since the relative position of each window in this plot shows the similarity between windows, 
this suggests that there are at least three extreme types of population structure 
typified by windows found in the three corners of the triangle,
and that other windows' population structure may be a mixture of those extremes.

\begin{figure}
    \begin{center}
       \includegraphics{Fig1_allchr_Together_MDS_plot_compact}
    \end{center}
    \caption{
        Variation in structure for windows across \textit{Drosophila melanogaster} chromosome arms.
        In all plots, each point represents one window along the genome.
         The first column shows the MDS visualization of relationships between windows,
         and the second and third columns show the position (midpoint) of each window along the genome
         against the first and second MDS coordinates, respectively. 
         Rows show results from different chromosome arms, from top to bottom these are 2L, 2R, 3L, 3R, and X. 
         Colors are consistent for plots in each row. 
         Vertical black lines show the breakpoints of known polymorphic inversions
         (from top to bottom: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
         \label{fig:mds_chr2L}
    }
\end{figure}

To investigate these extremes, for each chromosome arm we visually selected three ``extreme'' windows in the MDS plot,
and take out the 5\% of windows that are closest to it in the MDS coordinates;
highlighted these windows' positions along the genome (colors in Figure \ref{fig:mds_chr2L}),
and produced PCA plots for each collection of windows
(shown in Figure \ref{fig:pca_by_pop}).

\begin{figure}
    \begin{center}
       \includegraphics[width=1\textwidth]{Fig2_pca_plots_allchr_3peaks_label_update}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L},
        on each chromosome arm of \textit{Drosophila melanogaster}.
        In all plots, each point represents a sample. 
        The first column shows the combined PCA plot for windows 
        whose points are colored green in Figure \ref{fig:mds_chr2L}; 
        the second is for orange windows; 
        and third is for purple windows.
         \label{fig:pca_by_pop}
    }
\end{figure}

The variation in population structure turns out to be almost entirely explained by
several large inversions that are polymorphic in these samples, identified in \citet{corbett2012population}. 
Frequencies of each inversion in the sample are given in Table \ref{tab:inversion_freqs}.
\plr{Add this table.}
We recolored the PCA plots in Figure \ref{fig:pca_by_pop} by the orientation of the inversion for each sample,
obtaining Figure \ref{fig:pca_by_inversion}.
Taking chromosome arm 2L as an example,
the two regions of similar, extreme population structure
shown in green in the first row of Figure \ref{fig:mds_chr2L}
lie directly around the breakpoints of the iversion In(2L)t,
and the PCA plots in the first rows of figure \ref{fig:pca_by_inversion}
shows that population structure here is mostly determined by inversion orientation.
The regions shown in purple lie near the centromere,
and have population structure reflective of two axes of variation,
seen in figure \ref{fig:pca_by_pop},
which correspond roughly to latitude within Africa and to degree of cosmopolitan admixture respectively
(see \citet{pool2015natural} for more details).
\plr{check that reference}
The regions shown in orange mostly lie inside the inversion,
and show population structure that is a mixture between the other two,
as expected due to recombination within the (long) inversion \citep{kirkpatrick2015chromosome}.
\plr{check that reference too}
Similar results are found in other chromosome arms,
\plr{add chromosome X}
except that the situation of Chr3R is a little more complicated,
due to the coexistence of two polymorphic inversions (In(3R)K and In(3R)P) at intermediate frequency.
However, peaks in MDS coordinates along the chromosome show the breakpoints of both,
although we have not highlighted portions of the MDS plot that pick these out.
Two further inversions, In(3L)P and In(3R)Mo, are at very low frequency in these samples, 
and do not significantly influence population structure.

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the corresponding polymorphic inversions, 
         In(2L)t, In(2R)NS, In(3L)OK and In(3R)K respectively. 
         (Data from \citet{lack2015drosophila}.
        \label{fig:color_inver}
    }
\end{figure}

%%%%%%%%%%%
\subsection{Human}

As for the \textit{Drosophila} data, we ran our method separately on all 22 human autosomes, 
with results shown in Supplementary Figure XXX.
\plr{Add plots for remaining autosomes to supplement.}
On each, variation in population structure was dominated by a small number of windows
having similar population structure to each other that differed dramatically from the rest of the chromosome.
Such regions seem to often coincide with inversions:
for instance, the eleven windows that are outliers in the first MDS coordinate of chromosome 8 (Figure \ref{fig:mds_human}b) 
coincide with the position of known polymorphic inversions on 8p23. 
Similar results are found in other chromosomes that have known inversions,
shown in Figure \ref{fig:mds_human}.
PCA plots of these windows show a characteristic trimodal shape,
presumably distinguishing samples having each of the three diploid genotypes for each inversion orientation
(although we do not have data on orientation status).
PCA has been previously proposed as a method to computationally identify inversions \citep{ma2012investigation},
but understanding how well the method can distinguish inversions from regions of low recombination
is beyond the scope of this paper.

We also ran the method on all 22 autosomes together, 
and found that, remarkably, 
the inversion on chromosome 8 is still the most striking outlying signal.
Further investigation would likely benefit from a denser set of SNPs,
which would allow a finer genomic resolution.

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig4_POPRES_Together_MDS_plot_chr8_15_17}
    \end{center}
    \caption{
         Variation in structure between windows on human chromosomes 8, 15, and 17. 
         Each point in each plot represents a window.
         The first column shows the MDS visualization of relationships between windows;
         the second and third columns show the two MDS coordinates of each window against its position (midpoint) along the chromosome. 
         Rows, from top to bottom show chromosomes 8, 15, and 17. 
         The vertical red lines show the breakpoints of known inversions from \citep{antonacci2009characterization},
         whose values are listed in \plr{add these in a table}.
        \label{fig:mds_human}
    }
\end{figure}

@@

%%%%%%%%
\subsection*{Acknowledgements}

We are indebted to John Pool, Russ Corbett-Detig, Matilde Cordeiro, and Peter Chang 
for assistence with obtaining data and interpreting results
(especially inversion status of \textit{D.~melanogaster} samples).
Thanks also go to Yaniv Brandvain and Barbara Engelhardt for helpful comments 
and for encouraging the project.
